---
title: "Microbiome Analysis using Phyloseq"
output:
  html_document:
    df_print: paged
---

[Phyloseq R package](https://joey711.github.io/phyloseq/index.html)

[Phyloseq tutorial](https://vaulot.github.io/tutorials/Phyloseq_tutorial.html)

# Dataset

```{r}
rm(list=ls(all=TRUE))
library(tidyverse)
library(microbiome)
library(ggh4x)
```

Read relative abundance dataset:
```{r}
physeq <- read_rds("Data/phyloseq.rds")

physeq
```

Extract taxon table:
```{r}
tax_table <- physeq %>% tax_table() %>% as.data.frame() %>% rownames_to_column("OTU") %>% as_tibble()

tax_table
```

Sample attributes:
```{r}
sample_attributes <- physeq %>% sample_data() %>% data.frame() %>% rownames_to_column("Sample") %>% colnames()

sample_attributes
```

Experimental groups:
```{r}
physeq %>% sample_data() %>% as_tibble() %>% count(ATTRIBUTE_Study_Group)
```

# Relative family abundances

Convert Phyloseq object to long format:
```{r}
data_long <- physeq %>% 
  psmelt() %>% as_tibble()

data_long
```

What are the most abundant families across all samples?
```{r}
family_abundance <- data_long %>% 
  group_by(Family) %>% 
  summarize(Abundance = sum(Abundance)/n_distinct(Sample)) %>% 
  arrange(-Abundance)

family_abundance
```

* The most abundant families are by far Propionibacteriaceae and Staphylococcaceae

Select the largest families for downstream analysis:
```{r}
major_families <- family_abundance$Family[1:6]

major_families
```

## Family level

Add additional column Major_Families to the long dataset as well as to the Phyloseq object:
```{r}
data_long <- data_long %>% 
  mutate(
    Major_Families = if_else(
      Family %in% major_families,
      Family, "Other families"
    ),
    Major_Families = factor(Major_Families, levels = c(major_families, "Other families"))
  )

tax_table(physeq) <- tax_table(physeq) %>% 
  as.data.frame() %>% 
  mutate(
    Major_Families = if_else(
      Family %in% major_families,
      Family, "Other families"
    )
  ) %>% 
  as.matrix()
```

Plot relative abundance of the major families:
```{r}
data_long %>%
  inner_join(
    data_long %>% 
      group_by(ATTRIBUTE_Study_Group, Sample) %>% 
      summarize(
        Propionibacteriaceae = sum(Abundance[Family == "Propionibacteriaceae"]),
        .groups = "drop_last"
        ) %>% 
      arrange(-Propionibacteriaceae) %>% 
      mutate(
        Sample_No = row_number()
      ) %>% 
      select(ATTRIBUTE_Study_Group, Sample, Sample_No),
    by = c("ATTRIBUTE_Study_Group", "Sample")
  ) %>% 
  mutate(`Relative abundance` = Abundance) %>% 
  ggplot() +
  geom_col(
    aes(interaction(Sample_No, ATTRIBUTE_Study_Group), `Relative abundance`, fill = Major_Families)
    ) +
  scale_x_discrete(guide = guide_axis_nested(n.dodge = 0, title = "Study group")) +
  theme(
    axis.ticks.x = element_blank(),
    strip.text.x = element_text(angle = 90, vjust = 0.5, hjust=0),
    legend.position = "bottom",
    legend.title = element_blank()
    ) +
  scale_fill_brewer(palette = "Set1")
```

* Most volunteers from both study groups are dominated by either Propionibacteriaceae or Staphylococcaceae. Three volunteers have a major proportion of Polyomaviridae, a few more have a minor proportion.
* Although the range of relative proportions of Propionibacteriaceae and Staphylococcaceae is similar in both study groups, there are more volunteers with a high proportion of Propionibacteriaceae in the Oily Scalp group.

Label samples with more than 50% Propionibacteriaceae in the Phyloseq object:
```{r}
sample_data(physeq) <- sample_data(physeq) %>% 
  data.frame() %>% 
  rownames_to_column("Sample") %>% 
  inner_join(
    data_long %>% 
      group_by(Sample) %>% 
      summarize(
        Propionibacteriaceae = sum(Abundance[Family == "Propionibacteriaceae"]),
        Polyomaviridae       = sum(Abundance[Family == "Polyomaviridae"])
        ) %>% 
      mutate(
        Propionibacteriaceae = Propionibacteriaceae > 0.5,
        Polyomaviridae       = Polyomaviridae > 0.1
      ),
    by = "Sample"
  ) %>% 
  column_to_rownames("Sample")
```

Look at Propionibacteriaceae:
```{r}
data_long <- data_long %>% 
  mutate(
    Propionibacteria = if_else(
      Family == "Propionibacteriaceae" & is.na(Species), 
      Genus, Species
    )
  )

Propionibacteria <- data_long %>% 
  filter(Family == "Propionibacteriaceae") %>% 
  group_by(Propionibacteria) %>% 
  summarize(Abundance = sum(Abundance)/n_distinct(Sample)) %>% 
  arrange(-Abundance) %>% 
  pull(Propionibacteria)

data_long %>%
  inner_join(
    data_long %>% 
      group_by(ATTRIBUTE_Study_Group, Sample) %>% 
      summarize(
        Propionibacteriaceae = sum(Abundance[Family == "Propionibacteriaceae"]),
        .groups = "drop_last"
        ) %>% 
      arrange(-Propionibacteriaceae) %>% 
      mutate(
        Sample_No = row_number()
      ) %>% 
      select(ATTRIBUTE_Study_Group, Sample, Sample_No),
    by = c("ATTRIBUTE_Study_Group", "Sample")
  ) %>% 
  filter(Family == "Propionibacteriaceae") %>% 
  mutate(
    `Relative abundance` = Abundance
    ) %>% 
  ggplot() +
  geom_col(
    aes(interaction(Sample_No, ATTRIBUTE_Study_Group), `Relative abundance`, fill = Propionibacteria)
    ) +
  scale_x_discrete(guide = guide_axis_nested(n.dodge = 0, title = "Study group")) +
  theme(
    axis.ticks.x = element_blank(),
    strip.text.x = element_text(angle = 90, vjust = 0.5, hjust=0),
    legend.position = "bottom",
    legend.title = element_blank()
    ) +
  scale_fill_brewer(palette = "Set1", limits = Propionibacteria)
```

## Propionibacteriaceae

Look at strains of Propionibacteriaceae:
```{r}
data_long <- data_long %>% 
 mutate(
   Propionibacterium_varieties = if_else(
     Family == "Propionibacteriaceae" & is.na(Variety), 
     Propionibacteria, Variety
   )
 )

Propionibacterium_varieties <- data_long %>% 
  filter(Family == "Propionibacteriaceae") %>% 
  group_by(Propionibacterium_varieties) %>% 
  summarize(Abundance = sum(Abundance)/n_distinct(Sample)) %>% 
  arrange(-Abundance) %>% 
  pull(Propionibacterium_varieties)

data_long %>%
  inner_join(
    data_long %>% 
      group_by(ATTRIBUTE_Study_Group, Sample) %>% 
      summarize(
        Propionibacteriaceae = sum(Abundance[Family == "Propionibacteriaceae"]),
        .groups = "drop_last"
        ) %>% 
      arrange(-Propionibacteriaceae) %>% 
      mutate(
        Sample_No = row_number()
      ) %>% 
      select(ATTRIBUTE_Study_Group, Sample, Sample_No),
    by = c("ATTRIBUTE_Study_Group", "Sample")
  ) %>% 
  filter(Family == "Propionibacteriaceae") %>% 
  mutate(
    `Relative abundance` = Abundance
    ) %>% 
  ggplot() +
  geom_col(
    aes(interaction(Sample_No, ATTRIBUTE_Study_Group), `Relative abundance`, fill = Propionibacterium_varieties)
    ) +
  scale_x_discrete(guide = guide_axis_nested(n.dodge = 0, title = "Study group")) +
  theme(
    axis.ticks.x = element_blank(),
    strip.text.x = element_text(angle = 90, vjust = 0.5, hjust=0),
    legend.position = "right",
    legend.title = element_blank()
    ) +
  scale_fill_brewer(palette = "Set1", limits = Propionibacterium_varieties)
```

* The null hypothesis cannot be rejected that Propionibacterium_acnes_unclassified, Propionibacteriaceae_unclassified, and GCF_000221145 are in fact the same strain.

* All taxa from this family should be summarized as Propionibacteriaceae, with almost all reads being Propionibacterium_acnes

## Staphylococcaceae

Look at Staphylococcaceae:
```{r}
Staphylococci <- data_long %>% 
  filter(Family == "Staphylococcaceae") %>% 
  group_by(Species) %>% 
  summarize(Abundance = sum(Abundance)/n_distinct(Sample)) %>% 
  arrange(-Abundance) %>% 
  pull(Species) %>% 
  .[1:3]

data_long %>%
  inner_join(
    data_long %>% 
      group_by(ATTRIBUTE_Study_Group, Sample) %>% 
      summarize(
        Staphylococcaceae = sum(Abundance[Family == "Staphylococcaceae"]),
        .groups = "drop_last"
        ) %>% 
      arrange(-Staphylococcaceae) %>% 
      mutate(
        Sample_No = row_number()
      ) %>% 
      select(ATTRIBUTE_Study_Group, Sample, Sample_No),
    by = c("ATTRIBUTE_Study_Group", "Sample")
  ) %>% 
  filter(Family == "Staphylococcaceae") %>% 
  mutate(
    `Relative abundance` = Abundance,
    Staphylococci = if_else(Species %in% Staphylococci, Species, "Other Staphylococci")
    ) %>% 
  ggplot() +
  geom_col(
    aes(interaction(Sample_No, ATTRIBUTE_Study_Group), `Relative abundance`, fill = Staphylococci)
    ) +
  scale_x_discrete(guide = guide_axis_nested(n.dodge = 0, title = "Study group")) +
  theme(
    axis.ticks.x = element_blank(),
    strip.text.x = element_text(angle = 90, vjust = 0.5, hjust=0),
    legend.position = "right",
    legend.title = element_blank()
    ) +
  scale_fill_brewer(
    palette = "Set1",
    limits = Staphylococci %>% c("Other Staphylococci")
    )
```

Which taxa should be summarized into Staphylococcus_caprae_capitis?
```{r}
taxa_staph_caprae_capitis <- tax_table %>% 
  filter(Species == "Staphylococcus_caprae_capitis") %>% 
  pull(OTU)

tax_table %>% 
  filter(OTU %in% taxa_staph_caprae_capitis)
```

Which taxa should be summarized into Staphylococcus_epidermidis?
```{r}
taxa_staph_epidermidis <- tax_table %>% 
  filter(Species == "Staphylococcus_epidermidis") %>% 
  pull(OTU)

tax_table %>% 
  filter(OTU %in% taxa_staph_epidermidis)
```

Which taxa should be summarized into Staphylococcus_hominis?
```{r}
taxa_staph_hominis <- tax_table %>% 
  filter(Species == "Staphylococcus_hominis") %>% 
  pull(OTU)

tax_table %>% 
  filter(OTU %in% taxa_staph_hominis)
```

Which taxa should be summarized as other Staphylococci?
```{r}
taxa_staph_other <- tax_table %>% 
  filter(
    Family == "Staphylococcaceae" & 
      !OTU %in% c(taxa_staph_caprae_capitis, taxa_staph_epidermidis, taxa_staph_hominis)
    ) %>% 
  pull(OTU)

tax_table %>% 
  filter(OTU %in% taxa_staph_other)
```

## Polyomaviridae

Look at Polyomaviridae:
```{r}
Polyomaviri <- data_long %>% 
  filter(Family == "Polyomaviridae") %>% 
  group_by(Species) %>% 
  summarize(Abundance = sum(Abundance)/n_distinct(Sample)) %>% 
  arrange(-Abundance) %>% 
  pull(Species)

data_long %>%
  inner_join(
    data_long %>% 
      group_by(ATTRIBUTE_Study_Group, Sample) %>% 
      summarize(
        Polyomaviridae = sum(Abundance[Family == "Polyomaviridae"]),
        .groups = "drop_last"
        ) %>% 
      arrange(-Polyomaviridae) %>% 
      mutate(
        Sample_No = row_number()
      ) %>% 
      select(ATTRIBUTE_Study_Group, Sample, Sample_No),
    by = c("ATTRIBUTE_Study_Group", "Sample")
  ) %>% 
  filter(Family == "Polyomaviridae") %>% 
  mutate(`Relative abundance` = Abundance) %>% 
  ggplot() +
  geom_col(
    aes(interaction(Sample_No, ATTRIBUTE_Study_Group), `Relative abundance`, fill = Species)
    ) +
  scale_x_discrete(guide = guide_axis_nested(n.dodge = 0, title = "Study group")) +
  theme(
    axis.ticks.x = element_blank(),
    strip.text.x = element_text(angle = 90, vjust = 0.5, hjust=0),
    legend.position = "right",
    legend.title = element_blank()
    ) +
  scale_fill_brewer(palette = "Set1", limits = Polyomaviri)
```

Which taxa should be summarized into Polyomavirus_HPyV6?
```{r}
taxa_poly_hpyv6 <- tax_table %>% 
  filter(Species == "Polyomavirus_HPyV6") %>% 
  pull(OTU)

tax_table %>% 
  filter(OTU %in% taxa_poly_hpyv6)
```

Which taxa should be summarized into Polyomavirus_HPyV7
```{r}
taxa_poly_hpyv7 <- tax_table %>% 
  filter(Species == "Polyomavirus_HPyV7") %>% 
  pull(OTU)

tax_table %>% 
  filter(OTU %in% taxa_poly_hpyv7)
```

Which taxa should be summarized into Merkel_cell_polyomavirus?
```{r}
taxa_poly_merkel <- tax_table %>% 
  filter(Species == "Merkel_cell_polyomavirus") %>% 
  pull(OTU)

tax_table %>% 
  filter(OTU %in% taxa_poly_merkel)
```

## Skin paramters

Plot relative abundance of the major families, sorted by sebumeter score:
```{r}
data_long %>%
  inner_join(
    data_long %>% 
      group_by(ATTRIBUTE_Study_Group, Sample) %>% 
      summarize(
        Sebumeter_Score = Sebumeter_Score[1],
        .groups = "drop_last"
        ) %>% 
      arrange(Sebumeter_Score) %>% 
      mutate(
        Sample_No = row_number()
      ) %>% 
      select(ATTRIBUTE_Study_Group, Sample, Sample_No),
    by = c("ATTRIBUTE_Study_Group", "Sample")
  ) %>% 
  mutate(`Relative abundance` = Abundance) %>% 
  ggplot() +
  geom_col(
    aes(interaction(Sample_No, ATTRIBUTE_Study_Group), `Relative abundance`, fill = Major_Families)
    ) +
  scale_x_discrete(guide = guide_axis_nested(n.dodge = 0, title = "Study group")) +
  theme(
    axis.ticks.x = element_blank(),
    strip.text.x = element_text(angle = 90, vjust = 0.5, hjust=0),
    legend.position = "bottom",
    legend.title = element_blank()
    ) +
  scale_fill_brewer(palette = "Set1")

data_long %>% 
  group_by(ATTRIBUTE_Study_Group, Sample) %>% 
  summarize(
    Sebumeter_Score = Sebumeter_Score[1],
    .groups = "drop_last"
    ) %>% 
  arrange(Sebumeter_Score) %>% 
  mutate(
    Sample_No = row_number(),
    `Sebumeter score` = Sebumeter_Score
  ) %>% 
  ggplot() +
  geom_col(
    aes(interaction(Sample_No, ATTRIBUTE_Study_Group), `Sebumeter score`, fill = "")
    ) +
  scale_x_discrete(guide = guide_axis_nested(n.dodge = 0, title = "Study group")) +
  theme(
    axis.ticks.x = element_blank(),
    strip.text.x = element_text(angle = 90, vjust = 0.5, hjust=0),
    legend.position = "none",
    legend.title = element_blank()
    ) +
  scale_fill_brewer(palette = "Set1")
```

Plot relative abundance of the major families, sorted by Skicon score:
```{r}
data_long %>%
  inner_join(
    data_long %>% 
      group_by(Sample) %>% 
      summarize(
        Skicon_Score = Skicon_Score[1],
        .groups = "drop_last"
        ) %>% 
      arrange(Skicon_Score) %>% 
      mutate(
        Sample_No = row_number()
      ) %>% 
      select(Sample, Sample_No),
    by = "Sample"
  ) %>% 
  mutate(`Relative abundance` = Abundance) %>% 
  ggplot() +
  geom_col(
    aes(Sample_No, `Relative abundance`, fill = Major_Families)
    ) +
  theme(
    axis.ticks.x = element_blank(),
    axis.text.x = element_blank(),
    axis.title.x = element_blank(),
    legend.position = "bottom",
    legend.title = element_blank()
    ) +
  scale_fill_brewer(palette = "Set1")

data_long %>% 
  group_by(Sample) %>% 
  summarize(
    Skicon_Score = Skicon_Score[1],
    .groups = "drop_last"
    ) %>% 
  arrange(Skicon_Score) %>% 
  mutate(
    Sample_No = row_number(),
    `Skicon score` = Skicon_Score
  ) %>% 
  ggplot() +
  geom_col(
    aes(Sample_No, `Skicon score`, fill = "")
    ) +
  theme(
    axis.ticks.x = element_blank(),
    axis.text.x = element_blank(),
    axis.title.x = element_blank(),
    legend.position = "none",
    legend.title = element_blank()
    ) +
  scale_fill_brewer(palette = "Set1")
```

# Ordination analysis

Define function for axis labels:
```{r}
get_axis_labels <- function(x) str_c(
  "PC",
  1:length(x$values$Relative_eig),
  " [",
  (100 * x$values$Relative_eig) %>% round(1),
  "%]"
  )
```

## Bray-Curtis distance

Calculate ordination:
```{r}
ordination <- physeq %>% 
  ordinate("PCoA", "bray")
```

Plot samples:
```{r}
axis_labels <- get_axis_labels(ordination)

plot_ordination(
  physeq,
  ordination,
  type = "samples",
  color = "Dominat_Propioni"
  ) +
  scale_color_brewer(palette = "Set1", limits = c(T, F)) +
  labs(
    title = "PCoA, Bray-Curtis dissimilarity",
    x = axis_labels[1], y = axis_labels[2]
    )

plot_ordination(
  physeq,
  ordination,
  type = "samples",
  color = "ATTRIBUTE_Study_Group"
  ) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "PCoA, Bray-Curtis dissimilarity",
    x = axis_labels[1], y = axis_labels[2]
    )

plot_ordination(
  physeq,
  ordination,
  type = "samples",
  color = "ATTRIBUTE_Ethnic_Group"
  ) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "PCoA, Bray-Curtis dissimilarity",
    x = axis_labels[1], y = axis_labels[2]
    )

plot_ordination(
  physeq,
  ordination,
  type = "samples",
  color = "ATTRIBUTE_BiologicalSex"
  ) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "PCoA, Bray-Curtis dissimilarity",
    x = axis_labels[1], y = axis_labels[2]
    )

plot_ordination(
  physeq,
  ordination,
  type = "samples",
  color = "Sebumeter_Score"
  ) +
  scale_color_viridis_c() +
  labs(
    title = "PCoA, Bray-Curtis dissimilarity",
    x = axis_labels[1], y = axis_labels[2]
    )

plot_ordination(
  physeq,
  ordination,
  type = "samples",
  color = "Skicon_Score"
  ) +
  scale_color_viridis_c() +
  labs(
    title = "PCoA, Bray-Curtis dissimilarity",
    x = axis_labels[1], y = axis_labels[2]
    )

plot_ordination(
  physeq,
  ordination,
  type = "samples",
  color = "AgeInYears"
  ) +
  scale_color_viridis_c() +
  labs(
    title = "PCoA, Bray-Curtis dissimilarity",
    x = axis_labels[1], y = axis_labels[2]
    )

plot_ordination(
  physeq,
  ordination,
  type = "samples",
  color = "Propionibacteriaceae"
  ) +
  scale_color_brewer(palette = "Set1", limits = c(T, F)) +
  labs(
    title = "PCoA, Bray-Curtis dissimilarity",
    x = axis_labels[1], y = axis_labels[2]
    )

plot_ordination(
  physeq,
  ordination,
  type = "samples",
  color = "Polyomaviridae"
  ) +
  scale_color_brewer(palette = "Set1", limits = c(T, F)) +
  labs(
    title = "PCoA, Bray-Curtis dissimilarity",
    x = axis_labels[1], y = axis_labels[2]
    )
```

* There is no real separation into groups with oily or normal scalp.
* Samples with high sebumeter or skicon score are in the upper half of the plot, but samples with low sebumeter or skicon score are all around.

Plot families:
```{r}
plot_ordination(
  physeq,
  ordination,
  type = "taxa",
  color = "Major_Families"
  ) +
  scale_color_brewer(palette = "Set1", limits = c(major_families, "Other families"), name = NULL) +
  labs(
    title = "PCoA, Bray-Curtis dissimilarity",
    x = axis_labels[1], y = axis_labels[2],
    )
```

# Datasets for MMVec
## Reasonable taxon resolution

Aggregate dataset by major taxa. Check whether the sum of the abundances within each sample is 1 and whether no taxa were lost:
```{r}
data_major_taxa_long <- data_long %>% 
  mutate(
    Major_Taxa = case_when(
      OTU %in% taxa_poly_hpyv6  ~ "Polyomavirus HPyV6",
      OTU %in% taxa_poly_hpyv7  ~ "Polyomavirus HPyV7",
      OTU %in% taxa_poly_merkel ~ "Merkel Cell Polyomavirus",
      OTU %in% taxa_staph_caprae_capitis  ~ "Staphylococcus caprae or capitis",
      OTU %in% taxa_staph_epidermidis     ~ "Staphylococcus epidermidis",
      OTU %in% taxa_staph_hominis         ~ "Staphylococcus hominis",
      OTU %in% taxa_staph_other           ~ "Other Staphylococci",
      T ~ Major_Families
    ),
    Major_Taxa = Major_Taxa %>% factor(levels = c("Propionibacteriaceae", "Staphylococcus caprae or capitis", "Staphylococcus epidermidis", "Staphylococcus hominis", "Other Staphylococci", "Polyomavirus HPyV6", "Polyomavirus HPyV7", "Merkel Cell Polyomavirus", "Malasseziaceae", "Corynebacteriaceae", "Micrococcaceae", "Other families"))
  ) %>% 
  group_by(across(c(sample_attributes, Major_Families, Major_Taxa))) %>% 
  summarize(
    N_Taxa = n(),
    Abundance = sum(Abundance), 
    .groups = "drop"
    )

data_major_taxa_long %>% 
  group_by(Sample) %>% 
  summarize(Abundance = sum(Abundance), .groups = "drop")

data_major_taxa_long %>% 
  group_by(Major_Taxa) %>% 
  summarize(N_Taxa = N_Taxa[1]) %>% 
  summarize(N_Taxa = sum(N_Taxa))
```

Transform to wide format and save to file:
```{r}
data_major_taxa_wide <- data_major_taxa_long %>% 
  pivot_wider(id_cols = Major_Taxa, names_from = Sample, values_from = Abundance)

data_major_taxa_wide %>% write_tsv("Data/data_major_taxa_wide.tsv")

data_major_taxa_wide
```

Plot relative abundances:
```{r}
data_major_taxa_long %>%
  inner_join(
    data_major_taxa_long %>% 
      filter(Major_Taxa == "Propionibacteriaceae") %>% 
      arrange(-Abundance) %>% 
      group_by(ATTRIBUTE_Study_Group) %>% 
      mutate(
        Sample_No = row_number()
      ) %>% 
      select(ATTRIBUTE_Study_Group, Sample, Sample_No),
    by = c("ATTRIBUTE_Study_Group", "Sample")
  ) %>% 
  mutate(`Relative abundance` = Abundance) %>% 
  ggplot() +
  geom_col(
    aes(interaction(Sample_No, ATTRIBUTE_Study_Group), `Relative abundance`, fill = Major_Taxa)
    ) +
  scale_x_discrete(guide = guide_axis_nested(n.dodge = 0, title = "Study group")) +
  theme(
    axis.ticks.x = element_blank(),
    strip.text.x = element_text(angle = 90, vjust = 0.5, hjust=0),
    legend.position = "right",
    legend.title = element_blank()
    ) +
  scale_fill_brewer(palette = "Set3")
```

Plot relative abundances, sorted by sebumeter score:
```{r}
data_major_taxa_long %>%
  inner_join(
    data_major_taxa_long %>% 
      filter(Major_Taxa == "Propionibacteriaceae") %>% 
      arrange(Sebumeter_Score) %>% 
      group_by(ATTRIBUTE_Study_Group) %>% 
      mutate(
        Sample_No = row_number()
      ) %>% 
      select(ATTRIBUTE_Study_Group, Sample, Sample_No),
    by = c("ATTRIBUTE_Study_Group", "Sample")
  ) %>% 
  mutate(`Relative abundance` = Abundance) %>% 
  ggplot() +
  geom_col(
    aes(interaction(Sample_No, ATTRIBUTE_Study_Group), `Relative abundance`, fill = Major_Taxa)
    ) +
  scale_x_discrete(guide = guide_axis_nested(n.dodge = 0, title = "Study group")) +
  theme(
    axis.ticks.x = element_blank(),
    strip.text.x = element_text(angle = 90, vjust = 0.5, hjust=0),
    legend.position = "right",
    legend.title = element_blank()
    ) +
  scale_fill_brewer(palette = "Set3")

data_major_taxa_long %>% 
  group_by(ATTRIBUTE_Study_Group, Sample) %>% 
  summarize(
    Sebumeter_Score = Sebumeter_Score[1],
    .groups = "drop_last"
    ) %>% 
  arrange(Sebumeter_Score) %>% 
  mutate(
    Sample_No = row_number(),
    `Sebumeter score` = Sebumeter_Score
  ) %>% 
  ggplot() +
  geom_col(
    aes(interaction(Sample_No, ATTRIBUTE_Study_Group), `Sebumeter score`, fill = "")
    ) +
  scale_x_discrete(guide = guide_axis_nested(n.dodge = 0, title = "Study group")) +
  theme(
    axis.ticks.x = element_blank(),
    strip.text.x = element_text(angle = 90, vjust = 0.5, hjust=0),
    legend.position = "none",
    legend.title = element_blank()
    ) +
  scale_fill_brewer(palette = "Set1")
```

Plot relative abundances, sorted by sebumeter score:
```{r}
data_major_taxa_long %>%
  inner_join(
    data_major_taxa_long %>% 
      filter(Major_Taxa == "Propionibacteriaceae") %>% 
      arrange(Skicon_Score) %>% 
      mutate(
        Sample_No = row_number()
      ) %>% 
      select(ATTRIBUTE_Study_Group, Sample, Sample_No),
    by = c("ATTRIBUTE_Study_Group", "Sample")
  ) %>% 
  mutate(`Relative abundance` = Abundance) %>% 
  ggplot() +
  geom_col(
    aes(Sample_No, `Relative abundance`, fill = Major_Taxa)
    ) +
  theme(
    axis.ticks.x = element_blank(),
    axis.text.x = element_blank(),
    axis.title.x = element_blank(),
    legend.position = "right",
    legend.title = element_blank()
    ) +
  scale_fill_brewer(palette = "Set3")

data_major_taxa_long %>% 
  group_by(Sample) %>% 
  summarize(
    Skicon_Score = Skicon_Score[1],
    .groups = "drop_last"
    ) %>% 
  arrange(Skicon_Score) %>% 
  mutate(
    Sample_No = row_number(),
    `Skicon score` = Skicon_Score
  ) %>% 
  ggplot() +
  geom_col(
    aes(Sample_No, `Skicon score`, fill = "")
    ) +
  theme(
    axis.ticks.x = element_blank(),
    axis.text.x = element_blank(),
    axis.title.x = element_blank(),
    legend.position = "none",
    legend.title = element_blank()
    ) +
  scale_fill_brewer(palette = "Set1")
```

## Propionibacteriaceae and Staphylococcaceae,

Filter dataset for Propionibacteriaceae and Staphylococcaceae and re-scale the abundances within each sample to 1:
```{r}
data_propi_staph_long <- data_major_taxa_long %>% 
  filter(Major_Families %in% c("Propionibacteriaceae", "Staphylococcaceae")) %>% 
  group_by(Sample) %>% 
  mutate(Abundance = Abundance / sum(Abundance)) %>% 
  ungroup()

data_propi_staph_long %>% 
  group_by(Sample) %>% 
  summarize(Abundance = sum(Abundance), .groups = "drop")

data_propi_staph_long %>% 
  group_by(Major_Taxa) %>% 
  summarize(N_Taxa = N_Taxa[1]) %>% 
  summarize(N_Taxa = sum(N_Taxa))
```

Transform to wide format and save to file:
```{r}
data_propi_staph_wide <- data_propi_staph_long %>% 
  pivot_wider(id_cols = Major_Taxa, names_from = Sample, values_from = Abundance)

data_propi_staph_wide %>% write_tsv("Data/data_propi_staph_wide.tsv")

data_propi_staph_wide
```

Plot relative abundances:
```{r}
data_propi_staph_long %>%
  inner_join(
    data_propi_staph_long %>% 
      filter(Major_Taxa == "Propionibacteriaceae") %>% 
      arrange(-Abundance) %>% 
      group_by(ATTRIBUTE_Study_Group) %>% 
      mutate(
        Sample_No = row_number()
      ) %>% 
      select(ATTRIBUTE_Study_Group, Sample, Sample_No),
    by = c("ATTRIBUTE_Study_Group", "Sample")
  ) %>% 
  mutate(`Relative abundance` = Abundance) %>% 
  ggplot() +
  geom_col(
    aes(interaction(Sample_No, ATTRIBUTE_Study_Group), `Relative abundance`, fill = Major_Taxa)
    ) +
  scale_x_discrete(guide = guide_axis_nested(n.dodge = 0, title = "Study group")) +
  theme(
    axis.ticks.x = element_blank(),
    strip.text.x = element_text(angle = 90, vjust = 0.5, hjust=0),
    legend.position = "right",
    legend.title = element_blank()
    ) +
  scale_fill_brewer(palette = "Set3")
```

Plot relative abundances, sorted by sebumeter score:
```{r}
data_propi_staph_long %>%
  inner_join(
    data_propi_staph_long %>% 
      filter(Major_Taxa == "Propionibacteriaceae") %>% 
      arrange(Sebumeter_Score) %>% 
      group_by(ATTRIBUTE_Study_Group) %>% 
      mutate(
        Sample_No = row_number()
      ) %>% 
      select(ATTRIBUTE_Study_Group, Sample, Sample_No),
    by = c("ATTRIBUTE_Study_Group", "Sample")
  ) %>% 
  mutate(`Relative abundance` = Abundance) %>% 
  ggplot() +
  geom_col(
    aes(interaction(Sample_No, ATTRIBUTE_Study_Group), `Relative abundance`, fill = Major_Taxa)
    ) +
  scale_x_discrete(guide = guide_axis_nested(n.dodge = 0, title = "Study group")) +
  theme(
    axis.ticks.x = element_blank(),
    strip.text.x = element_text(angle = 90, vjust = 0.5, hjust=0),
    legend.position = "right",
    legend.title = element_blank()
    ) +
  scale_fill_brewer(palette = "Set3")

data_propi_staph_long %>% 
  group_by(ATTRIBUTE_Study_Group, Sample) %>% 
  summarize(
    Sebumeter_Score = Sebumeter_Score[1],
    .groups = "drop_last"
    ) %>% 
  arrange(Sebumeter_Score) %>% 
  mutate(
    Sample_No = row_number(),
    `Sebumeter score` = Sebumeter_Score
  ) %>% 
  ggplot() +
  geom_col(
    aes(interaction(Sample_No, ATTRIBUTE_Study_Group), `Sebumeter score`, fill = "")
    ) +
  scale_x_discrete(guide = guide_axis_nested(n.dodge = 0, title = "Study group")) +
  theme(
    axis.ticks.x = element_blank(),
    strip.text.x = element_text(angle = 90, vjust = 0.5, hjust=0),
    legend.position = "none",
    legend.title = element_blank()
    ) +
  scale_fill_brewer(palette = "Set1")
```

Plot relative abundances, sorted by sebumeter score:
```{r}
data_propi_staph_long %>%
  inner_join(
    data_propi_staph_long %>% 
      filter(Major_Taxa == "Propionibacteriaceae") %>% 
      arrange(Skicon_Score) %>% 
      mutate(
        Sample_No = row_number()
      ) %>% 
      select(ATTRIBUTE_Study_Group, Sample, Sample_No),
    by = c("ATTRIBUTE_Study_Group", "Sample")
  ) %>% 
  mutate(`Relative abundance` = Abundance) %>% 
  ggplot() +
  geom_col(
    aes(Sample_No, `Relative abundance`, fill = Major_Taxa)
    ) +
  theme(
    axis.ticks.x = element_blank(),
    axis.text.x = element_blank(),
    axis.title.x = element_blank(),
    legend.position = "right",
    legend.title = element_blank()
    ) +
  scale_fill_brewer(palette = "Set3")

data_propi_staph_long %>% 
  group_by(Sample) %>% 
  summarize(
    Skicon_Score = Skicon_Score[1],
    .groups = "drop_last"
    ) %>% 
  arrange(Skicon_Score) %>% 
  mutate(
    Sample_No = row_number(),
    `Skicon score` = Skicon_Score
  ) %>% 
  ggplot() +
  geom_col(
    aes(Sample_No, `Skicon score`, fill = "")
    ) +
  theme(
    axis.ticks.x = element_blank(),
    axis.text.x = element_blank(),
    axis.title.x = element_blank(),
    legend.position = "none",
    legend.title = element_blank()
    ) +
  scale_fill_brewer(palette = "Set1")
```

# Session Info

```{r}
timestamp()
sessionInfo()
```

