---
title: "Microbiome Dataset"
output:
  html_document:
    df_print: paged
---

```{r}
rm(list=ls(all=TRUE))
library(tidyverse)
library(phyloseq)
```

# Relative abundances
## Data input

Read relative abunce table (from Hunter):
```{r}
data_readcount <- read_tsv("Data/data_rel_abundance_baseline.tsv")
```

## OTU table

Assemble otu table. We do not need column TAXON_DESCRIPTION:
```{r}
otu_table <- data_readcount %>% 
  select(-TAXON_DESCRIPTION) %>% 
  column_to_rownames("TAXON_ID") %>% 
  phyloseq::otu_table(taxa_are_rows = T)

otu_table[1:6, 1:6]
```

Check if the sum of all taxa is 1, as expected for a relative abundance table:
```{r}
otu_table %>% 
  apply(2, sum)
```

# Sample metadata
## Data input

Read sample metadata:
```{r}
sample_metadata <- read_tsv("Data/sample_metadata_baseline.tsv", na = "#N/A")
```

## Sample table

Generate sample table:
```{r}
sample_table <- sample_metadata %>%
  column_to_rownames("External_Reference") %>% 
  phyloseq::sample_data()

sample_table %>% data.frame()
```

# Taxonomy
## Data input

The first two columns from the input data should be sufficient as taxonomy table:
```{r}
taxonomy <- data_readcount %>% 
  select(TAXON_ID, TAXON_DESCRIPTION)

taxonomy
```

## Taxonomy table

Generatare taxonomy table:
```{r}
tax_table <- taxonomy %>% 
  separate(
    TAXON_DESCRIPTION, sep = "; ", 
    into = c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species", "Variety"),
    extra = "drop", fill = "right"
    ) %>% 
  column_to_rownames("TAXON_ID") %>% 
  as.matrix() %>% 
  apply(c(1, 2), str_remove, "^[a-z]__") %>% 
  phyloseq::tax_table()

tax_table[1:6, ]
```

Check for missing taxon assignments:
```{r}
tax_table %>% apply(2, function(x) sum(is.na(x)))
```

Which taxon has missing phylum assignment?
```{r}
tax_table[tax_table[, "Phylum"] %>% is.na(), ]
```

Downstream analysis (relative abundance bar plots) suggests that this taxon is from one sample only, and that this sample comprises only this taxon:
```{r}
otu_table["f8c13bb601e2113b166ef926763585b4", ]
otu_table[, "EMR_04_N-22_RD"]
```

* This is correct. Taxon f8c13bb601e2113b166ef926763585b4 and sample EMR_04_N-22_RD need to be removed from the dataset. Will do this on the phyloseq object.

# Phyloseq dataset
## Phyloseq object

Generate Phyloseq object:
```{r}
physeq <- phyloseq::phyloseq(
  otu_table,
  tax_table,
  sample_table
)

physeq
```

* The phyloseq object has 56 samples, same as the relative abundance table
* The sample metadata is for 74 samples
* This is Ok. All samples in the sample metadata have metabolite data, but some are missing microbiome data.

Subset Phyloseq object to remove taxon f8c13bb601e2113b166ef926763585b4 and sample EMR_04_N-22_RD (see above):
```{r}
physeq <- physeq %>% 
  subset_samples(sample_names(physeq) != "EMR_04_N-22_RD") %>% 
  subset_taxa(taxa_names(physeq) != "f8c13bb601e2113b166ef926763585b4")
physeq
```

Write Phyloseq objects to file:
```{r}
physeq %>% write_rds("Data/phyloseq.rds")
```
