---
title: "GNPS Node Annotations"
output: html_notebook
---

```{r}
library(tidyverse)
library(wordcloud)

rm(list=ls(all=TRUE))
masses <- c(C = 12, H = 1.007825, N = 14.00307400443, O = 15.99491461957, P = 30.973761998, S = 31.9720711744)
```

# Data input

Read node annotations from file:
```{r}
node_annotations <- read_delim("mmvec skicon.csv")
```

Molecular networks:
```{r}
node_annotations$componentindex %>% sort() %>% unique()
```

Column names:
```{r}
node_annotations %>% colnames() %>% sort()
```

# Canopus

Canopus annotations:
```{r}
node_annotations %>% 
  select(starts_with("CAN_")) %>% 
  lapply(sort) %>% lapply(unique)
```

```{r}
word_freq <- tibble(
  Words = node_annotations$CAN_subclass
) %>% 
  filter(!is.na(Words)) %>% 
  mutate(Words = Words %>% str_remove_all("\"")) %>% 
  group_by(Words) %>% 
  summarize(Freq = n()) %>% 
  arrange(-Freq)

word_freq
```

```{r}
wordcloud(word_freq$Words, word_freq$Freq, min.freq = 1)
```

# Molecular formulas

SIRIUS/Zodiac annotations:
```{r}
node_annotations %>% 
  select(starts_with("CAN")) %>% 
  lapply(sort) %>% lapply(unique)
```

Molecular formula distribution
```{r}
parse_formula <- function(x) {
  if (is.na(x)) tibble() else {
    elements <- x %>% str_remove_all("[0-9]") %>% str_split("") %>% unlist()
    numbers  <- x %>% str_split("[A-Z]") %>% unlist %>% 
      .[2:length(.)] %>% if_else(. == "", "1", .) %>% as.integer()
    tibble(Elements = elements, Numbers = numbers)
  }
}

formula_distribution <- node_annotations %>% 
  mutate(
    Parsed_Formula = CAN_molecularFormula %>% map(parse_formula)
  ) %>% 
  select(Shared_Name = `shared name`, CAN_molecularFormula, Parsed_Formula) %>% 
  unnest(Parsed_Formula) %>% 
  pivot_wider(names_from = Elements, values_from = Numbers, values_fill = 0)

formula_distribution
```

Plot molecular formula distribution:
```{r}
formula_distribution %>% 
  pivot_longer(all_of(c("C", "H", "N", "O")), names_to = "Elements", values_to = "Numbers") %>% 
  ggplot() +
  geom_bar(aes(Numbers)) +
  facet_wrap("Elements", scales = "free")
```

Plot molecular formula distribution (C > 15):
```{r}
formula_distribution %>% 
  filter(C > 15) %>% 
  pivot_longer(all_of(c("C", "H", "N", "O")), names_to = "Elements", values_to = "Numbers") %>% 
  ggplot() +
  geom_bar(aes(Numbers)) +
  facet_wrap("Elements", scales = "free")
```

Plot molecular formula distribution (C > 15 & N == 2):
```{r}
formula_distribution %>% 
  filter(C > 15 & N == 2) %>% 
  pivot_longer(all_of(c("C", "H", "N", "O")), names_to = "Elements", values_to = "Numbers") %>% 
  ggplot() +
  geom_bar(aes(Numbers)) +
  facet_wrap("Elements", scales = "free")
```

Consensus formula (C > 15):
```{r}
consensus_formula <- formula_distribution %>% 
  filter(C > 15 & N == 2) %>% 
  pivot_longer(all_of(c("C", "H", "N", "O")), names_to = "Elements", values_to = "Numbers") %>% 
  group_by(Elements, Numbers) %>% 
  summarise(N = n(), .groups = "drop_last") %>% 
  arrange(-N) %>% 
  summarise(Numbers = Numbers[1])

consensus_formula
```

Print consonsus formula:
```{r}
consensus_formula %>% 
  mutate(Element_Number = str_c(Elements, Numbers)) %>% 
  pull(Element_Number) %>% 
  str_c(collapse = "")
```

Exact mass from consensus formula:
```{r}
consensus_formula %>% 
  mutate(Masses = masses[Elements] * Numbers) %>% 
  .$Masses %>% sum()
```

# Masses

Adducts:
```{r}
node_annotations %>% 
  ggplot() +
  geom_bar(aes(CAN_adduct))
```

Distribution of parent m/z:
```{r}
node_annotations %>% 
  ggplot() +
  geom_histogram(aes(`precursor mass`), bins = 50)
```

Distribution of CANOPUS masses:
```{r}
formula_distribution %>% 
  filter(C > 15 & N == 2) %>% 
  pivot_longer(all_of(c("C", "H", "N", "O")), names_to = "Elements", values_to = "Numbers") %>% 
  mutate(Masses = masses[Elements] * Numbers) %>% 
  group_by(Shared_Name) %>% 
  summarise(Masses = sum(Masses)) %>% 
  ggplot() +
  geom_histogram(aes(Masses), bins = 50)
```