---
title: "GNPS Node Annotations"
output: html_notebook
---

```{r}
library(tidyverse)
library(wordcloud)

rm(list=ls(all=TRUE))
masses <- c(C = 12, H = 1.007825, N = 14.00307400443, O = 15.99491461957, P = 30.973761998, S = 31.9720711744)
```

# Feature metadata

Read feature annotations from file:
```{r}
feature_info <- rbind(
  read_tsv("feature_metadata/C18neg_feature_metadata_consolidated_is_microbial.tsv", guess_max = 100000) %>% 
    mutate(FEATURE_ID = paste0("X95", formatC(`#featureID`,        width = 5, flag = "0", format = "d"))) %>% 
    mutate(FAMILY_ID   = paste0("X95", formatC(GNPS_componentindex, width = 4, flag = "0", format = "d"))),
  read_tsv("feature_metadata/C18pos_feature_metadata_consolidated_is_microbial.tsv", guess_max = 100000) %>% 
    mutate(FEATURE_ID = paste0("X94", formatC(`#featureID`,        width = 5, flag = "0", format = "d"))) %>% 
    mutate(FAMILY_ID   = paste0("X94", formatC(GNPS_componentindex, width = 4, flag = "0", format = "d"))),
  read_tsv("feature_metadata/HILICneg_feature_metadata_consolidated_is_microbial.tsv", guess_max = 100000) %>% 
    mutate(FEATURE_ID = paste0("X97", formatC(`#featureID`,        width = 5, flag = "0", format = "d"))) %>% 
    mutate(FAMILY_ID   = paste0("X97", formatC(GNPS_componentindex, width = 4, flag = "0", format = "d"))),
  read_tsv("feature_metadata/HILICpos_feature_metadata_consolidated_is_microbial.tsv", guess_max = 100000) %>% 
    mutate(FEATURE_ID = paste0("X96", formatC(`#featureID`,        width = 5, flag = "0", format = "d"))) %>% 
    mutate(FAMILY_ID   = paste0("X96", formatC(GNPS_componentindex, width = 4, flag = "0", format = "d")))
  ) %>% 
  mutate(FAMILY_ID = if_else(str_detect(FAMILY_ID, "-001$"), "Singleton", FAMILY_ID))
```

# Figure 3: GNPS
## Family X960074
### Canopus

Canopus annotations:
```{r}
current_family <- feature_info %>% 
  filter(FAMILY_ID == "X960074")

current_family %>% 
  select(starts_with("CAN_")) %>% 
  lapply(sort) %>% lapply(unique)
```

```{r}
word_freq <- tibble(
  Words = current_family$CAN_subclass
) %>% 
  filter(!is.na(Words)) %>% 
  mutate(Words = Words %>% str_remove_all("\"")) %>% 
  group_by(Words) %>% 
  summarize(Freq = n()) %>% 
  arrange(-Freq)

word_freq
```

```{r}
wordcloud(word_freq$Words, word_freq$Freq, min.freq = 1, scale = c(1.2, .4))
```

### Molecular formulas

SIRIUS/Zodiac annotations:
```{r}
current_family %>% 
  select(starts_with("SIR")) %>% 
  lapply(sort) %>% lapply(unique)
```

Molecular formula distribution
```{r}
parse_formula <- function(x) {
  if (is.na(x)) tibble() else {
    elements <- x %>% str_remove_all("[0-9]") %>% str_split("") %>% unlist()
    numbers  <- x %>% str_split("[A-Z]") %>% unlist %>% 
      .[2:length(.)] %>% if_else(. == "", "1", .) %>% as.integer()
    tibble(Elements = elements, Numbers = numbers)
  }
}

formula_distribution <- current_family %>% 
  mutate(
    Parsed_Formula = SIR_MF_Zod_molecularFormula %>% map(parse_formula)
  ) %>% 
  select(FEATURE_ID, SIR_MF_Zod_molecularFormula, Parsed_Formula) %>% 
  unnest(Parsed_Formula) %>% 
  pivot_wider(names_from = Elements, values_from = Numbers, values_fill = 0)

formula_distribution
```

Plot molecular formula distribution:
```{r}
formula_distribution %>% 
  pivot_longer(3:ncol(.), names_to = "Elements", values_to = "Numbers") %>% 
  ggplot() +
  geom_bar(aes(Numbers)) +
  facet_wrap("Elements", scales = "free")
```

Plot molecular formula distribution (N %in% c(1, 3)):
```{r}
formula_distribution %>% 
  filter(N %in% c(1, 3)) %>% 
  pivot_longer(3:ncol(.), names_to = "Elements", values_to = "Numbers") %>% 
  ggplot() +
  geom_bar(aes(Numbers)) +
  facet_wrap("Elements", scales = "free")
```

Plot molecular formula distribution (N == 1):
```{r}
formula_distribution %>% 
  filter(N == 1) %>% 
  pivot_longer(3:ncol(.), names_to = "Elements", values_to = "Numbers") %>% 
  ggplot() +
  geom_bar(aes(Numbers)) +
  facet_wrap("Elements", scales = "free")
```

Consensus formula (N == 1):
```{r}
consensus_formula_n1 <- formula_distribution %>% 
  filter(N == 1) %>% 
  pivot_longer(3:ncol(.), names_to = "Elements", values_to = "Numbers") %>% 
  group_by(Elements, Numbers) %>% 
  summarise(N = n(), .groups = "drop_last") %>% 
  arrange(-N) %>% 
  summarise(Numbers = Numbers[1])

consensus_formula_n1
```

Exact mass from consensus formula:
```{r}
consensus_formula_n1 %>% 
  mutate(Masses = masses[Elements] * Numbers) %>% 
  .$Masses %>% sum()
```

Plot molecular formula distribution (N == 3):
```{r}
formula_distribution %>% 
  filter(N == 3) %>% 
  pivot_longer(3:ncol(.), names_to = "Elements", values_to = "Numbers") %>% 
  ggplot() +
  geom_bar(aes(Numbers)) +
  facet_wrap("Elements", scales = "free")
```

Consensus formula (N == 3):
```{r}
consensus_formula_n3 <- formula_distribution %>% 
  filter(N == 3) %>% 
  pivot_longer(3:ncol(.), names_to = "Elements", values_to = "Numbers") %>% 
  group_by(Elements, Numbers) %>% 
  summarise(N = n(), .groups = "drop_last") %>% 
  arrange(-N) %>% 
  summarise(Numbers = Numbers[1])

consensus_formula_n3
```

Exact mass from consensus formula:
```{r}
consensus_formula_n3 %>% 
  mutate(Masses = masses[Elements] * Numbers) %>% 
  .$Masses %>% sum()
```

### Masses

Distribution of parent masses:
```{r}
current_family %>% 
  ggplot() +
  geom_histogram(aes(`GNPS_precursor mass`), bins = 50)
```

Distribution of SIRIUS masses:
```{r}
formula_distribution %>% 
  filter(N %in% c(1, 3)) %>% 
  pivot_longer(3:ncol(.), names_to = "Elements", values_to = "Numbers") %>% 
  mutate(Masses = masses[Elements] * Numbers) %>% 
  group_by(FEATURE_ID) %>% 
  summarise(Masses = sum(Masses)) %>% 
  ggplot() +
  geom_histogram(aes(Masses), bins = 50)
```

# Figure 4: MMvec
## Family X950225
### Canopus

Canopus annotations:
```{r}
current_family <- feature_info %>% 
  filter(FAMILY_ID == "X950225")

current_family %>% 
  select(starts_with("CAN_")) %>% 
  lapply(sort) %>% lapply(unique)
```

```{r}
word_freq <- tibble(
  Words = current_family$CAN_subclass
) %>% 
  filter(!is.na(Words)) %>% 
  mutate(Words = Words %>% str_remove_all("\"")) %>% 
  group_by(Words) %>% 
  summarize(Freq = n()) %>% 
  arrange(-Freq)

word_freq
```

```{r}
wordcloud(word_freq$Words, word_freq$Freq, min.freq = 1, scale = c(1.2, .4))
```

## Family X950396
### Canopus

Canopus annotations:
```{r}
current_family <- feature_info %>% 
  filter(FAMILY_ID == "X950396")

current_family %>% 
  select(starts_with("CAN_")) %>% 
  lapply(sort) %>% lapply(unique)
```

```{r}
word_freq <- tibble(
  Words = current_family$CAN_subclass
) %>% 
  filter(!is.na(Words)) %>% 
  mutate(Words = Words %>% str_remove_all("\"")) %>% 
  group_by(Words) %>% 
  summarize(Freq = n()) %>% 
  arrange(-Freq)

word_freq
```

```{r}
wordcloud(word_freq$Words, word_freq$Freq, min.freq = 1, scale = c(1.4, .2))
```

## Family X960242
### Canopus

Canopus annotations:
```{r}
current_family <- feature_info %>% 
  filter(FAMILY_ID == "X960242")

current_family %>% 
  select(starts_with("CAN_")) %>% 
  lapply(sort) %>% lapply(unique)
```

```{r}
word_freq <- tibble(
  Words = current_family$CAN_subclass
) %>% 
  filter(!is.na(Words)) %>% 
  mutate(Words = Words %>% str_remove_all("\"")) %>% 
  group_by(Words) %>% 
  summarize(Freq = n()) %>% 
  arrange(-Freq)

word_freq
```

```{r}
wordcloud(word_freq$Words, word_freq$Freq, min.freq = 1, scale = c(1.2, .2))
```
